---
title: "DiffExp analysis for Kelava et al. 2020"
author: "Alex Kalinka"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: hide
    keep_md: yes
    toc: yes
    depth: 3
    highlight: tango
    number_sections: yes
    theme: spacelab
    toc_float:
      collapsed: yes
    df_print: paged
---

# Setup

The `delboy` R package lives here:

https://github.com/alextkalinka/delboy

```{r}
options(warn=-1)
suppressMessages(library(knitr))
suppressMessages(library(delboy))
suppressMessages(library(dplyr))
suppressMessages(library(tidyr))
suppressMessages(library(ggplot2))


if(packageVersion("delboy") != "0.0.1")
  stop("this analysis is designed to work with delboy v0.0.1")

# Batch-corrected data pre-computed (non-parametric batch correction in sva::ComBat is very slow).
data.bcorr.day35 <- readRDS("../data/day35-R-objects/bcorr.non-param--day35.rds")
data.bcorr.day35_hormone <- readRDS("../data/day35-R-objects/bcorr.non-param--day35-hormone.rds")

```

# Overview

We start from batch-corrected data as this step is time-consuming. See the Appendix for the steps taken to produce the batch-corrected data from the `salmon` quants.

# Differential Expression analysis

The low-expression `filter_cutoff` is set to 0 because a filter of 40 (average of 4 TPM/sample) was applied when the batch-correction was pre-computed.

```{r}
suppressMessages(
  suppressWarnings(
    dboy.day35 <- delboy::run_delboy(
      data = data.bcorr.day35,
      group_1 = colnames(data.bcorr.day35)[c(4,8,12)],
      group_2 = colnames(data.bcorr.day35)[c(2:3,5,7,9,11)],
      normalize = NULL,
      filter_cutoff = 0,
      gene_column = "gene_id",
      bcorr_data_validation = data.bcorr.day35_hormone)
  )
)

```

## Visualizing hits

Below we visualize the relative expression of samples within genes to show that up- and down-regulated genes fall into opposite corners of the plot, as expected.

```{r}
suppressMessages(
data.rel <- data.bcorr.day35 %>%
  select(-ends_with("E")) %>%
  gather(key = "sample", value = "abundance", -gene_id) %>%
  rowwise() %>%
  mutate(Treatment = tail(unlist(strsplit(sample,"_")),1),
         Treatment = ifelse(Treatment == "DHT","T",Treatment),
         Treatment = ifelse(Treatment=="ctrl","Control","Testosterone/DHT"),
         batch = paste("b",unlist(strsplit(sample,"_"))[2],sep="")) %>%
  ungroup() %>%
  group_by(gene_id) %>%
  filter(sum(abundance) > 40 & abundance > 0) %>%
  mutate(min_sb = (abundance-min(abundance)),
         TPM_relative = min_sb/max(min_sb)) %>%
  summarise(TPM_relative.Control = mean(TPM_relative[Treatment=="Control"]),
            TPM_relative.Androgen = mean(TPM_relative[Treatment=="Testosterone/DHT"])) %>%
  ungroup() %>%
  mutate(Gene = case_when(gene_id %in% dboy.day35$elnet_results$genes.up ~ "Hit_Upregulated",
                          gene_id %in% dboy.day35$elnet_results$genes.down ~ "Hit_Downregulated",
                          TRUE ~ "Not called"))
)

# Plot.
data.rel %>%
  ggplot(aes(TPM_relative.Control, TPM_relative.Androgen, color = Gene)) +
  geom_bin2d() +
  geom_point(data=data.rel %>% filter(Gene=="Hit_Upregulated"),
             aes(TPM_relative.Control, TPM_relative.Androgen, color="Hit_Upregulated")) +
  geom_point(data=data.rel %>% filter(Gene=="Hit_Downregulated"),
             aes(TPM_relative.Control, TPM_relative.Androgen, color="Hit_Downregulated")) +
  ggtitle("TPM (normalised within genes): Up- and down-regulated hits")

```

## Performance

`delboy` has a 10-fold higher sensitivity than `DESeq2` (adjusted p-value < 0.1) and with a modest FDR of 4% (3 false positives).

```{r}
kable(dboy.day35$performance_stats_corr_FP)

plot(dboy.day35, type = "fc_expr", ylim=c(0,1.5), xlim=c(0.5,4))

plot(dboy.day35, type = "fc_expr_FN", ylim=c(0,1.5), xlim=c(0.5,4))

```

## Hit list

```{r}
db.hits <- delboy::hits(dboy.day35)
cat(paste("Number of hits:",nrow(db.hits)),"\n",sum(db.hits$log2FoldChange > 0),"up-regulated\n",sum(db.hits$log2FoldChange < 0),"down-regulated\n\n")
kable(db.hits)

```

# Appendix

To run the code in the following section, change the `eval=F` option in the code-chunk header to `eval=T`.

## Batch correcting data from `salmon` quants

Batch 502 and 802 E looked like large outliers in the original sample PCA and were excluded from downstream analyses. To explore the total set of data, use the following path for the `path_samples_file` argument in the `read_salmon_quants` function below:

* "../data/day35-salmon-quants/samples.txt".

```{r, eval=F}
suppressMessages(library(tximport))
suppressMessages(library(tximportData))
suppressMessages(library(readr))
suppressMessages(library(DESeq2))
suppressMessages(library(sva))


read_salmon_quants <- function(path_samples_file, filter_cutoff = 40){
  # Tx2gene.
  tx_dir <- system.file("extdata", package="tximportData")
  tx2gene <- readr::read_csv(file.path(tx_dir, "tx2gene.gencode.v27.csv"))

  # Read in salmon quants for day35 data.
  path_quants <- "../data/day35-salmon-quants"
  samples <- read.table(path_samples_file, header=T)

  squant_files <- list.files(path_quants, recursive = T, full.names = T, pattern = "quant.sf")
  names(squant_files) <- sapply(squant_files, function(x) 
    gsub("^(.*)$", "\\1", unlist(strsplit(x,"/"))[4]))
  squant_files <- squant_files[match(samples$run, names(squant_files))]
  
  # Load salmon quants.
  txi <- tximport(squant_files, type="salmon", tx2gene=tx2gene)
  dds <- DESeqDataSetFromTximport(txi, colData = samples, 
                                  design = ~ batch + hormone)
  dds <- dds[rowSums(counts(dds)) > filter_cutoff,]
  var_st <- DESeq2::vst(dds, blind=FALSE)
  
  ret <- list(
    samples = samples,
    squant_files = squant_files,
    txi = txi,
    var_st = var_st,
    dds = dds,
    pca = plotPCA(var_st, intgroup=c("batch","hormone"), returnData=T)
  )
  return(ret)
}

# Prep data for batch correction by 'ComBat'.
prep_combat <- function(data, thresh){
  # 'thresh' usually set to 40.
  genes <- rownames(data)
  batch <- sapply(colnames(data), function(x) unlist(strsplit(x,"_"))[2])
  names(batch) <- NULL
  data <- data.frame(gene = genes, data) %>%
    filter(rowSums(.[2:ncol(.)]) > thresh)
  return(list(data = data, batch_names = batch))
}

quants <- read_salmon_quants("../data/day35-salmon-quants/samples-d_t_grouped.txt")

# Sample PCA prior to batch correction.
ggplot(quants$pca, aes(PC1, PC2, shape=hormone, color=batch)) + 
  geom_point(size=3) +
  ggtitle("Sample PCA: pre-batch-correction")

# ComBat non-parametric Batch-correction.
data.combat.day35 <- prep_combat(quants$txi$abundance, thresh=40)
data.bcorr <- sva::ComBat(as.matrix(data.combat.day35$data[,2:ncol(data.combat.day35$data)]),
                          data.combat.day35$batch_names, par.prior = F)

as.data.frame(prcomp(t(data.bcorr))$x) %>%
  mutate(sample = colnames(data.bcorr)[1:ncol(data.bcorr)]) %>%
  rowwise() %>%
  mutate(hormone = unlist(strsplit(sample,"_"))[4], batch = unlist(strsplit(sample,"_"))[2]) %>%
  ungroup() %>%
  ggplot(aes(PC1, PC2, shape = hormone, color = batch)) +
  geom_point(size = 3) +
  ggtitle("Sample PCA: post-batch-correction")

```



